---
title: "ISSA Skin Spectrum Reflectance"
author: "Junhui He"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  workflowr::wflow_html:
    toc: yes
    number_sections: yes
editor_options:
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# Load required libraries
library(tidyverse)
library(knitr)
library(scales)
library(patchwork)
library(ggbeeswarm)
library(ggridges)
library(ggmulti)
library(paletteer)
library(stringr)

library(gridExtra)
library(openxlsx)

# Set up the project root directory
F = rprojroot::is_rstudio_project$make_fix_file()

# Get available palettes for potential use
palettedf <- tibble(paletteer::palettes_d_names)

# Set global chunk options for the document
knitr::opts_chunk$set(
  echo = FALSE,        # Don't show code in output
  include = TRUE,      # Include output
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Don't show warnings
  message = FALSE,     # Don't show messages
  fig.retina = 2,      # Higher resolution figures
  fig.width = 8,       # Default figure width
  fig.height = 6,      # Default figure height
  out.width = "100%"   # Full width in HTML output
)
```

# Introduction to the ISSA Dataset

The International Skin Spectra Archive (ISSA) offers a detailed collection of spectral and colorimetric data for human skin, encompassing 15,256 records from 2,113 subjects. This data spans from 2012 to 2024 and originates from eleven datasets curated by international laboratories across eight countries: the UK, Spain, China, Japan, Pakistan, Thailand, Iraq, and Saudi Arabia. Each dataset follows a standardised measurement protocol to maintain data consistency.

In the ISSA dataset, individual records provide extensive details including record number, data origin, subject identification, and skin type—categorised by ethnicity, gender, age, and body location. The dataset also includes detailed information on the measurement instruments used, such as type, specular component inclusion, wavelength range and interval.

Alongside spectral data, each sample also contains CIE colorimetric data, including tristimulus values, xy chromaticity coordinates, CIELAB parameters, etc., based on the CIE 1931 standard colorimetric observer and the CIE standard illuminant D65.

## Data Records

The datasheet arranges data across columns labelled A to BQ:

-   A: Unique record identifier
-   B: Data origin
-   C: Subject number
-   D to G: Ethnicity, gender, age group, and body location
-   H to L: Instrument details including type and spectral measurement specifics
-   N to BD: Spectral data from 360 nm to 780 nm
-   BF to BQ: CIE colorimetric data

## Skin Type

-   **Ethnicity**: CA (Caucasian), CN (Chinese), SA (South Asian), AF (African), IQ (Iraqi), TH (Thai), JP (Japanese), AB (Arabian)
-   **Gender**: F (Female), M (Male)
-   **Body Location**: 1 (Back of Hand), 2 (Cheek), 3 (Cheek bone), 4 (Chin), 5 (Ear Lobe), 6 (Forehead), 7 (Inner arm), 8 (Neck), 9 (Nose tip), 10 (Outer arm), 11 (Palm), 12 (Ring finger)

```{r load_issa}
# Load the ISSA skin spectrum reflectance data
issa_path <- "G:/Phd/Colloborator/Tina Lasisi/Meta/ISSA_17_Jan_2025_Yan_Lu.xlsx"
issa_code <- read.xlsx(issa_path, sheet = 1)
issa <- read.xlsx(issa_path, sheet = 2)

# Extract the skin spectrum data
issa_spectrum <- issa[10:nrow(issa), ] # 360 nm: 13rd column, 780 nm: 55th column
colnames(issa_spectrum) <- issa[9, ]
```

```{r util_functions}
# ==============================================================================
# DATA PROCESSING FUNCTIONS
# ==============================================================================

#' Calculate melanin index and prepare data for plotting
#' 
#' This function:
#' 1. Calculates the melanin index (CMI) using wavelengths 640-670nm
#' 2. Filters data to reasonable CMI range (20-130)
#' 3. Samples data to prevent plot overload
#' 4. Converts to long format for ggplot
#' 
#' @param data Processed spectral data frame
#' @param sample_size Number of spectra to sample (default 400)
#' @return Long-format data frame with id, wavelength, reflectance, and cmi
calculateMIndex <- function(data, sample_size = 400) {
  # Calculate melanin index using the formula from the literature
  # This uses the red region of the spectrum (640-670nm)
  data_processed <- data %>% 
    mutate(
      # Average reflectance in the red region
      eqn1 = ((`650` + `660` + (0.5 * `640`) + (0.5 * `660`)) / 3) / 100,
      # Convert to melanin index
      cmi = 100 * log10(1 / eqn1)
    )
  
  # Prepare data for plotting
  data_long <- data_processed %>%
    #mutate(id = row_number()) %>%
    # Filter to reasonable melanin index range
    filter(cmi >= 20 & cmi <= 130) %>% 
    # Sample to prevent plot overload
    # slice_sample(n = sample_size) %>% 
    # Convert to long format
    pivot_longer(cols = matches("^[0-9]+$"), 
                 names_to = "wavelength", 
                 values_to = "reflectance") %>% 
    # Extract numeric wavelength values
    mutate(wavelength = as.numeric(str_remove(wavelength, "\\D+"))) %>% 
    # Focus on visible spectrum
    filter(wavelength >= 400 & wavelength <= 700) # %>%
    # select(id, everything())
  
  return(data_long)
}

# ==============================================================================
# PLOTTING FUNCTIONS
# ==============================================================================

#' Generate skin spectra plot
#' 
#' Creates a plot showing reflectance spectra colored by melanin index
#' 
#' @param data Long-format data with wavelength, reflectance and cmi
#' @return A ggplot object
PlotSkinSpectra <- function(data, title = "Spectra from ISSA dataset") {
  plot <- ggplot(data, aes(x = wavelength, y = reflectance, 
                           group = C, color = cmi)) +
    geom_line(alpha = 0.6) +  # Add transparency to handle overlapping lines
    # Color gradient from dark to light skin
    scale_color_gradientn(
      colours = c("#42230c", "#a15c33", "#fbf7ec"),
      values = rescale(x = c(50, 40, 20)),
      oob = squish,
      limits = c(20, 130)
    ) +
    expand_limits(y = 0) +
    theme_classic() +
    labs(
      x = "Wavelength (nm)",
      y = "Percent Reflectance of Skin",
      color = "Melanin Index"
    ) +
    guides(color = guide_colorbar(reverse = TRUE)) +
    # Highlight the region used for melanin index calculation
    annotate("rect", xmin = 640, xmax = 670, ymin = 0, ymax = 70,
             alpha = 0.3, fill = "#ff0000") +
    labs(
      title = title #,
      # subtitle = "Red region (640-670nm) is used to calculate melanin index"
    )
  
  return(plot)
}

#' Wrapper function to process file and generate plot
#' 
#' @param file_name Path to the CSV file
#' @return A ggplot object
PlotSkinSpectraWrapper <- function(data, title, ethnicity = "CA", location = "6", gender = NULL) {
  # Process the data
  if(is.null(gender)) {
    processed_data <- data %>% filter(D == ethnicity, G == location)
  } else {
    processed_data <- data %>% filter(D == ethnicity, E == gender, G == location)
  }
  
  # Calculate melanin index
  data_with_mindex <- calculateMIndex(processed_data)
  
  # Generate the plot
  plot <- PlotSkinSpectra(data_with_mindex, title = title)
  
  return(plot)
}
```

# Plot Skin Spectrum Reflectance from ISSA

In this section, we present skin reflectance spectra (reflectance intensity vs. wavelength) at the forehead, comparing different ethnicities and genders.

```{r issa_spectrum_plot, fig.cap="ISSA skin reflectance spectra colored by melanin index", fig.height=5}
eths = c("CA", "CN", "SA", "AF", "IQ", "TH", "JP", "AB")
eths_full = c("Caucasians", "Chineses", "South Asians", "Africans", "Iraqis", "Thais", "Japaneses", "Arabians")

for (i in seq_along(eths)) {
  ethnicity = eths[i]
  ethnicity_full = eths_full[i]
  plot_female = PlotSkinSpectraWrapper(issa_spectrum, title = paste("Female", ethnicity_full,"at Forehead"), ethnicity = ethnicity, gender = "F", location = "6")
  plot_male = PlotSkinSpectraWrapper(issa_spectrum, title = paste("Male", ethnicity_full,"at Forehead"), ethnicity = ethnicity, gender = "M", location = "6")
  print(plot_female + plot_male + plot_layout(guides = "collect") & theme(legend.position = "bottom"))
}
```

# Construct Confidence Intervals

In this section, we construct confidence intervals of the skin reflectance spectra in ISSA. The intervals are obtained using the following methods:

-   **Naive Pointwise Confidence Intervals**

-   **Bootstrap Methods**

For compactness, we restrict our analysis to three ethnic groups---Caucasian, Chinese, and African---combining male and female participants, and focusing on the forehead location (site 6).

```{r ci_data, echo=TRUE}
# Extract the reflectance spectra of Caucasians, Chinese, and Africans at forehead
location = "6"
issa_spectrum_ca <- as.matrix(issa_spectrum %>% filter(D == "CA", G == location) %>% select("400":"700"))
issa_spectrum_cn <- as.matrix(issa_spectrum %>% filter(D == "CN", G == location) %>% select("400":"700"))
issa_spectrum_af <- as.matrix(issa_spectrum %>% filter(D == "AF", G == location) %>% select("400":"700"))
```

## Naive Pointwise Confidence Intervals

Naive Pointwise Confidence Intervals calculate the mean reflectance and the empirical quantiles (*e.g.*, $\alpha/2$ and $1-\alpha/2$ quantiles) at each wavelength. The pointwise $1-\alpha$ confidence interval is then constructed as: $$(\alpha/2 ~ \text{quantile}, ~ 1-\alpha/2 ~ \text{quantile} ).$$

```{r issa_spectrum_ci_naive, fig.cap="Naive pointwise confidence intervals for forehead reflectance spectra", fig.height=10, echo=TRUE}
NaivePointwiseCI <- function(data, level = 0.95) {
  wavelength = as.numeric(colnames(data))
  lower = apply(data, 2, quantile, probs = (1-level)/2)
  upper = apply(data, 2, quantile, probs = level + (1-level)/2)
  mean_curve = colMeans(data)
  return(data.frame(wavelength=wavelength, lower=lower, upper=upper, mean=mean_curve))
}

PlotPointwiseCI <- function(df_ci, title) {
  plot = ggplot(df_ci, aes(x = wavelength)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#fbf7ec") +
    geom_line(aes(y = mean), color = "#a15c33", linewidth = 1) +
    expand_limits(y = 0) +
    theme_classic() +
    labs(
      x = "Wavelength (nm)",
      y = "Percent Reflectance of Skin",
      color = "Melanin Index"
    ) +
    labs(
      title = title
    )
  return(plot)
}

# Calculate lower, upper bounds and mean values
issa_ci_ca = NaivePointwiseCI(issa_spectrum_ca, level = 0.95)
issa_ci_cn = NaivePointwiseCI(issa_spectrum_cn, level = 0.95)
issa_ci_af = NaivePointwiseCI(issa_spectrum_af, level = 0.95)

plot_ci_ca = PlotPointwiseCI(issa_ci_ca, title = "Naive Pointwise Confidence Intervals")
plot_ca = PlotSkinSpectraWrapper(issa_spectrum, title = "Caucasians at Forehead", ethnicity = "CA", location = 6)

plot_ci_cn = PlotPointwiseCI(issa_ci_cn, title = "Naive Pointwise Confidence Intervals")
plot_cn = PlotSkinSpectraWrapper(issa_spectrum, title = "Chinese at Forehead", ethnicity = "CN", location = 6)

plot_ci_af = PlotPointwiseCI(issa_ci_af, title = "Naive Pointwise Confidence Intervals")
plot_af = PlotSkinSpectraWrapper(issa_spectrum, title = "Africans at Forehead", ethnicity = "AF", location = 6)

print(((plot_ca | plot_ci_ca) / (plot_cn | plot_ci_cn) / (plot_af | plot_ci_af)))
```

## Bootstrap Confidence Intervals for the Mean Function

To calculate confidence intervals for the mean function, we can bootstrap:

1.  Resample $n$ curves with replacement.
2.  Compute the mean function for each resample.
3.  For each wavelength, compute the quantiles of the bootstrapped means.

```{r ci_bootstrap, fig.cap="Bootstrap pointwise confidence intervals for the mean function", fig.height=10, echo=TRUE}
BootstrapPointwiseCI <- function(data, level = 0.95) {
  n = nrow(data)
  m = ncol(data)
  B = 1000 # Bootstrap sample size
  
  boot_means = array(NA, c(B, m))
  for (b in 1:B) {
    idx = sample(1:n, n, replace = TRUE)
    boot_means[b, ] = colMeans(data[idx, ])
  }
  
  wavelength = as.numeric(colnames(data))
  lower = apply(boot_means, 2, quantile, probs = (1-level)/2)
  upper = apply(boot_means, 2, quantile, probs = level + (1-level)/2)
  mean_curve = colMeans(data)
  return(data.frame(wavelength=wavelength, lower=lower, upper=upper, mean=mean_curve))
}

# Calculate lower, upper bounds and mean values
issa_ci_ca_boot = BootstrapPointwiseCI(issa_spectrum_ca, level = 0.95)
issa_ci_cn_boot = BootstrapPointwiseCI(issa_spectrum_cn, level = 0.95)
issa_ci_af_boot = BootstrapPointwiseCI(issa_spectrum_af, level = 0.95)

plot_ci_ca_boot = PlotPointwiseCI(issa_ci_ca_boot, title = "Bootstrap Pointwise Confidence Intervals")
plot_ci_cn_boot = PlotPointwiseCI(issa_ci_cn_boot, title = "Bootstrap Pointwise Confidence Intervals")
plot_ci_af_boot = PlotPointwiseCI(issa_ci_af_boot, title = "Bootstrap Pointwise Confidence Intervals")
print(((plot_ci_ca | plot_ci_ca_boot) / (plot_ci_cn | plot_ci_cn_boot) / (plot_ci_af | plot_ci_af_boot)))
```

## Functional PCA based confidence intervals

Functional Principal Component Analysis is a classical statistical method for functional data analysis. Using this method, a random function is represented in the eigenbasis, which is an orthonormal basis of the Hilbert space that consists of the eigenfunctions of the autocovariance operator. 

For a squared-integrable stochastic process $\{X_t,~t\in T\}$, let 
$$\mu_t = \mathbb{E}X_t,$$
and 
$$G(s,t)=\mathbb{E}[(X_s-\mu_s)(X_t-\mu_t)]=\sum_{k=1}^\infty \lambda_k \varphi_k(s)\varphi_k(t),$$
where $\lambda_1\geq\lambda_2 \geq \cdots \geq 0$ are the eigenvalues and $\varphi_1,\varphi_2,\ldots$ are the eigenfunctions of a  linear Hilbert--Schmidt operator. By the Karhunen–Loève theorem, one can express the centered process in the eigenbasis,
$$X_t = \mu_t + \sum_{k=1}^\infty \xi_k \varphi_k(t),$$
where $\xi_k$ is endowed with the properties
$$\mathbb{E}\xi_k = 0, \quad \text{Var}(\xi_k)=\lambda_k,\quad \mathbb{E}\xi_k\xi_l = 0,~k\neq l.$$

```{r ci_fpca, fig.cap="FPCA-based pointwise confidence intervals for forehead reflectance spectra", fig.height=10, echo=TRUE}
# X: n x m matrix of curves (rows = curves, cols = grid points)
FPCAPointwiseCI <- function(X,
                            level = 0.95, ev_thresh = 0.99, K = NULL) {
  stopifnot(is.matrix(X))
  n <- nrow(X); m <- ncol(X)
  alpha <- 1 - level

  # 1) Estimate mean curve
  mu_hat <- colMeans(X)

  # 2) Centered data and covariance across t
  Xc <- sweep(X, 2, mu_hat, "-")
  C_hat <- crossprod(Xc) / (n - 1)   # m x m

  # 3) FPCA
  eig <- eigen(C_hat, symmetric = TRUE)
  lam <- pmax(eig$values, 0)
  Phi <- eig$vectors               # columns are eigenfunctions on the grid

  # choose K
  if (is.null(K)) {
    cumvar <- cumsum(lam) / sum(lam)
    K <- which(cumvar >= ev_thresh)[1]
  }
  lamK <- lam[1:K]; PhiK <- Phi[, 1:K, drop = FALSE]

  # 4) Pointwise SE for mean via diag(C)/n
  var_mu <- colSums( t(PhiK^2) * lamK )  # diag approx: sum_k lam_k * phi_k(t)^2 / n
  se_mu  <- sqrt(var_mu)

  # 5) Pointwise band
  z <- qnorm(1 - alpha/2)
  lower_pt <- mu_hat - z * se_mu
  upper_pt <- mu_hat + z * se_mu

  wavelength = as.numeric(colnames(X))
  
  list(
    wavelength = wavelength,
    mean = mu_hat,
    se = se_mu,
    lower = lower_pt,
    upper = upper_pt,
    lambda = lamK,
    phi = PhiK,
    K = K,
    ev_explained = sum(lamK) / sum(lam)
  )
}

# Calculate lower, upper bounds and mean values
issa_ci_ca_fpca_full = FPCAPointwiseCI(issa_spectrum_ca, level = 0.95)
issa_ci_ca_fpca = data.frame(wavelength = issa_ci_ca_fpca_full$wavelength,
                             lower = issa_ci_ca_fpca_full$lower,
                             upper = issa_ci_ca_fpca_full$upper,
                             mean = issa_ci_ca_fpca_full$mean)
issa_ci_cn_fpca_full = FPCAPointwiseCI(issa_spectrum_cn, level = 0.95)
issa_ci_cn_fpca = data.frame(wavelength = issa_ci_cn_fpca_full$wavelength,
                             lower = issa_ci_cn_fpca_full$lower,
                             upper = issa_ci_cn_fpca_full$upper,
                             mean = issa_ci_cn_fpca_full$mean)
issa_ci_af_fpca_full = FPCAPointwiseCI(issa_spectrum_af, level = 0.95)
issa_ci_af_fpca = data.frame(wavelength = issa_ci_af_fpca_full$wavelength,
                             lower = issa_ci_af_fpca_full$lower,
                             upper = issa_ci_af_fpca_full$upper,
                             mean = issa_ci_af_fpca_full$mean)

plot_ci_ca_fpca = PlotPointwiseCI(issa_ci_ca_fpca, title = "FPCA-based Pointwise Confidence Intervals")
plot_ci_cn_fpca = PlotPointwiseCI(issa_ci_cn_fpca, title = "FPCA-based Pointwise Confidence Intervals")
plot_ci_af_fpca = PlotPointwiseCI(issa_ci_af_fpca, title = "FPCA-based Pointwise Confidence Intervals")

print(((plot_ca | plot_ci_ca_fpca) / (plot_cn | plot_ci_cn_fpca) / (plot_af | plot_ci_af_fpca)))
```




